---
title: "Retail Transactions:  Monthly Report"
author: "Daniel Dasgupta"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r seed}
set.seed(41)
```

```{r libraries, echo = FALSE, warning=FALSE, message=FALSE}
library(data.table)
library(DT)
library(rmarkdown)
library(lubridate)

```

```{r constants}

#files constants
file.customers.data <- "customers.csv"
file.product.data <- "products.csv"
file.views.data <- "views -- January 2020.csv"
file.trans.data <- "transactions -- January 2020.csv"

#variables 
customer.id.name <- "customer_id"
age.name <- "age"
gender.name <- "gender"
income.name <- "income"
region.name <- "region"
category.name <- "category"
price.name <- "price"
product.id.name <- "product_id"

#thresholds
trans.threshold = 1
income.threshold = 10000

```

```{r functions}
round.numerics <- function(x, digits) {
  if (is.numeric(x)) {
    x <- round(x = x, digits = digits)
  }
  return(x)
}



rounded.sum <- function(x, digits){
  if(is.numeric(x) == TRUE){
    x <- round(x = sum(x), digits = digits)
  }
  return(x)
}


summarize.variable <- function(dat, variable.name, by.name ) {
  require(data.table)
  require(DT)
  setDT(dat)
  
  tab = dat[, .(mean = mean(get(variable.name), na.rm = T),
                sd = sd(get(variable.name), na.rm = T), 
                median = median(get(variable.name))), by = eval(by.name)]
  
  datatable(data = tab[, lapply(X = .SD, FUN = "round.numerics", digits = 1)], 
            rownames = F)
}
```

```{r load_data}
dat.customers <- fread(input = file.customers.data)
dat.products <- fread(input = file.product.data)
dat.views <- fread(input = file.views.data)
dat.trans <- fread(input = file.trans.data)

```



## Part 2:  Specific Questions {.tabset}

This part of the report will be directed to product managers throughout the client's company.  The idea is to give them the useful information they need to act on the specific questions they posed.  Plan your communication accordingly.

For this report, make sure to use all of the data that was provided to you for the month.  If you do note any issues with the data (Part 3), this can be reported to the engineering team for them to resolve.

### Q1

#### Question

During the first week of the month, what were the 10 most viewed products?  Show the results in a table with the product's identifier, category, and count of the number of views.

#### Answer

```{r part2_q1}


#clean time, separate date and hours/mins/sec by character T
dat.views$time <- gsub("T.*","",dat.views$time)

dat.views.first.week <-  dat.views[time >= "2020-01-01" & time <= "2020-01-07"]
tab.views.most.viewed.products <- dat.views.first.week[, .N, by = product.id.name]
setorderv(x = tab.views.most.viewed.products, cols = "N", order = -1)
tab.views.most.viewed.products <- tab.views.most.viewed.products[1:10]


#remove price from products table 
products = dat.products[, category, product_id]
part2_q1 <- merge(tab.views.most.viewed.products, products, by = product.id.name)
datatable(part2_q1)
```

### Q2

#### Question

During the whole month, what were the 10 most viewed products for each category?  Show the results in separate tables by category.  Including only the product's identifier and the count of the number of views.


#### Answer {.tabset}

```{r part2_q2_overall_calculations}

#filter the month dates 
jan.views <- dat.views[time <= '2020-01-31']

#count of most popular products 
tab.monthly.products <- jan.views[, .N, by = product.id.name]
#use products from previous question
part2_q2 <- merge(tab.monthly.products, products, by = product.id.name)
```

##### Shirt

```{r part2_q2_shirt}
shirt <- part2_q2[category == "shirt"]
setorderv(x = shirt, cols = "N", order= -1)
shirt <- shirt[1:10][,  N, product_id]
datatable(shirt)

```

##### Pants

```{r part2_q2_pants}

pants <- part2_q2[category == "pants"]
setorderv(x = pants, cols = "N", order= -1)
pants <- pants[1:10][,  N, product_id]
datatable(pants)
```

##### Coat

```{r part2_q2_coat}

coat <- part2_q2[category == "coat"]
setorderv(x = coat, cols = "N", order= -1)
coat <- coat[1:10][,  N, product_id]
datatable(coat)
```

##### Shoes

```{r part2_q2_shoes}

shoes <- part2_q2[category == "shoes"]
setorderv(x = shoes, cols = "N", order= -1)
shoes <- shoes[1:10][,  N, product_id]
datatable(shoes)
```

##### Hat

```{r part2_q2_hat}

hat <- part2_q2[category == "hat"]
setorderv(x = hat, cols = "N", order= -1)
hat <- hat[1:10][,  N, product_id]
datatable(hat)
```


### Q3

#### Question

What was the total revenue for each category of product during the month?  Show the results in a single table sorted in decreasing order.

#### Answer

```{r part2_q3}

#multiple quant by price 
dat.trans[, trans_revenue := price*quantity ]

#merge products with trans to get category

revenue.by.category <- merge(dat.trans, dat.products, by = c("product_id", "price"))
revenue.by.category <- revenue.by.category[, category, trans_revenue]

#calculate the average of each numeric variable grouped by specific
tab.revenue.by.category = revenue.by.category[, lapply(X = .SD, FUN = "rounded.sum", digits = 1), by = category.name]
setorderv(x = tab.revenue.by.category, cols = "trans_revenue", order = -1 )
datatable(tab.revenue.by.category)
```

### Q4

#### Question

Among customers with at least one transaction, show the average, median, and standard deviation of the customers' monthly spending on the site.

#### Answer

```{r part2_q4, warning = FALSE, echo=FALSE, results=FALSE}

#first get customers with more than one transaction
#create a transaction column 
dat.trans[, total_transactions := .N, keyby = customer.id.name]
dat.trans[total_transactions >= trans.threshold]

#summarize.variable(dat.trans, "trans_revenue", "customer_id")
```

```{r part2_q4_output, warning = FALSE}
summarize.variable(dat.trans, "trans_revenue", "customer_id")

```

### Q5

#### Question

What is the percentage distribution of spending by gender?  Show the amount of revenue and the percentage.

#### Answer

```{r part2_q5}

tab.spend.by.gender <- merge(dat.trans, dat.customers, by = customer.id.name)

totalrev <- round.numerics(sum(tab.spend.by.gender$price*tab.spend.by.gender$quantity), 2)
part2_q5 <-tab.spend.by.gender[, .(spending = sum(price*quantity), percent_dist = round((sum(price*quantity)/totalrev)*100, 2)), by = gender]
part2_q5$spending <- round.numerics(part2_q5$spending, 2)

datatable(part2_q5)
```

### Q6

#### Question

Using linear regression, what is the effect of an extra ten thousand dollars of income on monthly spending for a customer while adjusting for age, gender, and region?

#### Answer

```{r part2_q6}

dat.trans.top.spenders <- merge(dat.trans, dat.customers, by = customer.id.name)


dat.trans.top.spenders <- dat.trans[,.(amount_spent=sum(price)), by = customer_id]

dat.trans.top.spenders.total <- merge(dat.trans.top.spenders, dat.customers, by = customer.id.name)

mod1 <- lm(amount_spent ~ age + gender + region + income, dat = dat.trans.top.spenders.total)

part2_q6 <- summary(mod1)$coefficients[7,1]*income.threshold

part2_q6 <- round.numerics(part2_q6, 2)


```

An extra 10,000 dollars in income increases monthly spending for a customer by $`r part2_q6`, while adjusting for age, gender, and region.  

### Q7

#### Question

Among customers who viewed at least 1 product, how many had at least one purchase during the month?  Show the total number and as a percentage of the users with a view.

#### Answer

```{r part2_q7}

jan.views <- dat.views[time <= '2020-01-31']


dat.trans$time <- strftime(dat.trans$time, format = "%Y-%m-%d")
jan.trans <- dat.trans[time <= '2020-01-31']
jan.views$trans <- as.character(jan.views$trans)


trans_views <- merge(jan.views, jan.trans, by = c("customer_id", "product_id", "time"))

views_count <- jan.views[,.N, by = customer_id]

views_and_buy <- trans_views[,.N, by= customer_id]

percentage <- round(nrow(views_and_buy)/nrow(views_count)*100,2)

q7.tab <- matrix(c(nrow(views_count), nrow(views_and_buy),"100", percentage), ncol = 2, byrow =FALSE)
colnames(q7.tab) <- c('Count', 'Percentage')
rownames(q7.tab) <- c(">= 1 View", "View and Bought")
#as.table(q7.tab)

datatable(q7.tab)
```

### Q8

#### Question

Now let's look at the viewing habits in different age groups, including 18-34, 35-49, 50-64, and 65+.  Within each group, what were the mean, median, and standard deviation for the number of unique products viewed per customer?

#### Answer

```{r part2_q8}

unique.products <- dat.views[, .(count = length(unique(product_id))), by = customer_id]

views_customers <- merge(unique.products, dat.customers, by = "customer_id")

views_customers$age_group <- cut(views_customers$age, breaks = c(17,34,49,64,67), labels = c("18-34", "35-49", "50-64", "65+"))

q8_tab <- views_customers[, .(mean = mean(count), median = median(count), sd = sd(count)), by = age_group]

q8_tab$mean <-round(q8_tab$mean, 2)
q8_tab$sd <- round(q8_tab$sd, 2)

datatable(q8_tab)
```


### Q9 

#### Question

What is the correlation between a user's total page views and total spending?  For customers without a transaction, include their spending as zero.

#### Answer

```{r part2_q9}

view.count <- dat.views[, .(viewcount = .N), by = customer_id]

trans.total <- dat.trans[, .(spent = sum(price*quantity)), by = customer_id]

q9_dat <-merge(view.count, trans.total, by = "customer_id", all.x = TRUE)

q9_dat[is.na(q9_dat)] <- 0

corr <- round(cor(q9_dat$viewcount, q9_dat$spent)*100,2)
```
The correlation between a user's total page views and total spending is `r corr`. 

### Q10

#### Question

Which customer purchased the largest number of coats?  In the event of a tie, include all of the users who reached this value.  Show their identifiers and total volume.

#### Answer

```{r part2_q10}

q10 <- merge(dat.customers, dat.trans, by = customer.id.name)

q10 <- merge(q10, dat.products, by = c(product.id.name, "price"))


q10<- q10[, .(customer_id, product_id, category, trans_revenue, age, gender, income, region, quantity) ]

coats <- q10[category =="coat", .(bought = sum(quantity)), by = "customer_id"]
setorderv(coats, cols= "bought", order = -1)
#max(coats$bought)
coats <- coats[bought == 50]

datatable(coats)
```

